\huge \textbf{Parameter Forecasts From CMB Lensing and Galaxy Lensing Power- and Bispectra}

### 2025-04-04 paper/old/paper_old_version/paper_old_vers.tex
paper changes

% !TEX root = /Users/jonasfrugte/Desktop/Research_Project/paper/paper.tex
\documentclass[11pt]{article} % Choose a document class, like 'article', 'report', or 'revtex4-2' for scientific papers

% Packages to enhance functionality
\usepackage[utf8]{inputenc}   % Allows UTF-8 character encoding
\usepackage{amsmath}          % For advanced math typesetting
\usepackage{amssymb}          % Additional math symbols
\usepackage{graphicx}         % For including images
\usepackage{hyperref}         % For hyperlinks in the document
\usepackage{geometry}         % For setting up page geometry
\geometry{margin=1in}         % 1-inch margins all around
\usepackage{authblk}          % For author affiliations
\usepackage{siunitx}          % For consistent typesetting of units
\usepackage{subcaption}       % For subfigures within a figure
\usepackage{float}            % To control float positioning (e.g., figures/tables)
\usepackage{enumitem}         % Better control over list formatting
\usepackage{tocloft}
\usepackage{tikz}
\usepackage{tikz-3dplot}
\usetikzlibrary{shapes.geometric, arrows, backgrounds}
\usepackage{empheq}
\usepackage{booktabs}
\usepackage{authblk}

@@

\usepackage[backend=biber, style=numeric]{biblatex}
\addbibresource{sources.bib}

@@

\title{\huge \textbf{Parameter Forecasts From CMB Lensing and Galaxy Lensing Power- and Bispectra}}

\author[1]{\textbf{Jonas Frugte}}
\author[2]{\textbf{P. Daniel Meerburg}}

\affil[1]{Independent Researcher, The Netherlands}
\affil[2]{Van Swinderen Institute for Particle Physics and Gravity, University of Groningen, Nijenborgh 4, 9747 AG Groningen, The Netherlands}
\affil[ ]{}
\affil[ ]{\textit{E-mail:} \texttt{jonasfrugte@gmail.com}}

\date{20 February 2025} % Removes the default date

\begin{document}
\maketitle

% \textbf{
%     \huge \noindent 
%     Parameter Forecasts From CMB $\times$ Galaxy Lensing Powerspectra $\times$ Bispectra
%     \vspace{0.5cm}
%     \\
%     \normalsize
%     Jonas Frugte, Daan Meerburg\\
%     20 February 2025
%     }\\
% \vspace{0.5cm}
\hrule

\begin{abstract}
    (Placeholder text written with chatgpt) We present a Fisher forecast analysis of the constraints on cosmological parameters from measurements of CMB lensing and galaxy weak lensing power- and bispectra. Using simulated data consistent with upcoming surveys, we explore the sensitivity of lensing observables to parameters including the Hubble constant \(H_0\), baryon density \(\Omega_b h^2\), dark matter density \(\Omega_c h^2\), scalar spectral index \(n_s\), amplitude of scalar fluctuations \(A_s\), and the sum of neutrino masses \(\sum m_\nu\).

    Our results indicate that combining power- and bispectra improves parameter constraints compared to power spectra alone, reducing uncertainties by a factor of \(\sim X\%\). In particular, we forecast constraints of \(\sigma(H_0) = X.X\) km/s/Mpc, \(\sigma(\Omega_b h^2) = X.XXX\), \(\sigma(\Omega_c h^2) = X.XXX\), \(\sigma(n_s) = X.XXXX\), \(\sigma(A_s) = X.XX \times 10^{-9}\), and \(\sigma(\sum m_\nu) = X.XX\) eV. These results demonstrate the potential of bispectrum measurements in future surveys such as the Simons Observatory, CMB-S4, LSST, and Euclid.

    Additionally, we investigate the impact of different noise models and survey redshift distributions, finding that improvements in galaxy redshift accuracy significantly enhance constraints on \(\sum m_\nu\). We also discuss the degeneracies between parameters and assess their impact using eigenvalue analysis of the Fisher matrix.

    This study highlights the advantages of combining lensing power- and bispectra for cosmological parameter estimation and provides a roadmap for optimizing future surveys to maximize their scientific return.
\end{abstract}
\pagebreak
%\tableofcontents

% \section{Conventions}
% Use $\mathbf k$ for wave numbers (Fourier space), $l$ for multipole moments, $m$ for magnetic quantum number. $P^{XY}(k, \eta, \eta')$ and $B^{XYZ}(k_1,k_2,k_3, \eta_1, \eta_2, \eta_3)$ correspond to bi and power spectra of fields $X$, $Y$, and $Z$ in Fourier space. $P_l^{XY}(\eta,\eta')$ and $B_{l_1l_2l_3}^{XYZ}(\eta_1, \eta_2, \eta_3)$ correspond to the multipole moments of the bi and power spectra of fields $X$, $Y$, and $Z$. If all times are the same we may instead write only one time argument and if the time is current time then we omit the time argument entirely. In other words,
% $$
% P_l^{XY}(\eta):=P_l^{XY}(\eta,\eta), \quad P_l^{XY}:=P_l^{XY}(\eta_{\text{current}}), \quad \text{similarly for all other spectra}.
% $$
% If both fields are the same, we may only denote one field, e.g. 
% $$
% P_l^{X}(\eta) := P_l^{XX}(\eta).
% $$

% Our Fourier convention is
% \begin{align}
%     &\text{Forward Transform:} & \quad \tilde{f}(k) &= \int_{-\infty}^{\infty} f(x) \, e^{-i k x} \, dx\\
%     &\text{Inverse Transform:} & \quad f(x) &= \frac{1}{2\pi} \int_{-\infty}^{\infty} \tilde{f}(k) \, e^{i k x} \, dk
% \end{align}
% we choose this convention because it is in line with the convention used in the CAMB source code. We work in natural units where $c=1$.

@@

Weak lensing has been used as a probe to constrain cosmological parameters since the 2000's \cite{kilbinger2015cosmology,Bacon2000,Kaiser2000,Waerbeke2000,Bartelmann2001}. A range of upcoming surveys, such as the survey by the Simons Observatory (SO) \cite{Ade2019}, the Legacy Survey of Space and Time (LSST) \cite{Ivezic2019}, and Euclid \cite{Laureijs2011} aim to measure weak lensing of the CMB and of galaxies. They will do so much more accurately than previous surveys, enabling tighter constraints on parameters and progressing our understanding of physics as a whole. Typically, the main value of interest is the power spectrum (or equivalently, two point function) of the lensing potential. This can be measured in terms of the lensing convergence $\kappa$, lensing shear $\gamma$, or lensing potential $\psi$. In the weak lensing regime they all contain the same amount of information and can be easily converted into one another, see section \ref{sec:weaklensstats}. For the CMB the lensing potential is generally measured directly \cite{Planck2018Lensing}, while with galaxies one instead measures the shear, as this can be deduced from the ellipiticity distribution of the observed galaxies \cite{HoekstraJain2008}.

Even with Gaussian initial conditions, there are significant amounts of non-Gaussianity in the weak lensing signal, especially in galaxy lensing due to it being a probe of the more recent distribution of matter \cite{Bernardeau1997,Takada2003}. In addition to looking at the power spectrum, a natural next step is thus to measure the lensing bispectrum. This has already been done for signals from galaxies \cite{vanWaerbeke2002}, but no proper detection has been done of the CMB lensing bispectrum. Upcoming surveys such as the ones mentioned earlier may detect this signal. Therefore it is relevant to know the detectability of, and parameter constraints possible from, all 4 signals: CMB lensing power- and bispectra and galaxy lensing power- and bispectra. Notably, this includes looking at cross correlation as well.

In this article we aim to answer this question in the context of experimental parameters similar to those of next generation surveys such as mentioned earlier. We are especially interested in seeing if approximate parameter degeneracies can be broken by combining all data. Calculating the covariance matrix of unbiased optimal estimators of the cosmological parameters can be done by combining cosmological simulations (in our case we use the CAMB package \cite{Lewis2000}) with a Fisher matrix analysis. See figure \ref{fig:flowchart} for an overview of the calculation done in this paper. The parameters looked at are the Hubble constant, $H_0$, physical baryon density, $\Omega_b h^2$, physical cold dark matter density, $\Omega_c h^2$, scalar spectral index, $n_s$, amplitude of primordial scalar fluctuations, $A_s$, sum of neutrino masses $\sum m_\nu$, and dark energy equation of state parameter, $w_0$. Though, we will find that some of the parameters are poorly constrained even when combining all 4 signals and thus omit them from the analysis in the results. As a secondary purpose, this paper also includes a derivation for the Fisher matrix formalism in the non-trivial context of bispectra with multiple tracers. The final formula can already be found in the literature (see e.g.\ the appendix of \cite{Kalaja_2021}), but the authors of this paper were not able to find a derivation and hence included it.

The structure of this paper is as follows: we first review the theory of weak lensing and weak lensing statistics in sections \ref{sec:weaklensing} and \ref{sec:weaklensstats}. We then provide technical details of the calculation in section \ref{sec:calcdetails}, such as choices for noise and numerical approximations made in the code. In section \ref{sec:results} we present the results, consisting of signal to noise ratios and parameter constraints for different combinations of the CMB and galaxy lensing power- and bispectra. We conclude in section \ref{sec:discussion}. Appendix \ref{sec:fisher} contains a short review of the Fisher matrix formalism and gives a technical derivation of the formula used to calculate each Fisher matrix element in the case of bispectra measurements with multiple tracers. Finally, appendix \ref{sec:shear} explains how to relate the shear to the lensing potential in spherical harmonic space.

All code written for this project is uploaded on github at LINK .

\begin{figure}[h!]
    \centering
    \input{figures/calc_flowchart.tikz}
    \caption{Flowchart of steps involved in calculating the minimum error/uncertainty in cosmological parameter estimates using convergence and shear power- and bi-spectra.}
    \label{fig:flowchart}
\end{figure}

\section{Weak Lensing}\label{sec:weaklensing}
\subsection{Approximations Made / Regime}
\subsection{Perturbed Photon Paths}
Work in \textbf{conformal newtonian gauge}. In this section we'll work in units where $c=1$. Denoting conformal time and conformal radial distance by $\eta$ and $\chi$, respectively, the perturbed line element is given by
\begin{equation}
    \d s^2 = a^2(\eta)((1 + 2\Psi_N) \d \eta^2 - (1 + 2\Phi_N) \gamma_{ij}\d x^i \d x^j) 
\end{equation}

@@

In the unperturbed there are incoming radial solutions of the form $\chi = \eta_0 - \eta$. We can express $\hat\lambda$ in terms of $\eta$ using The 0-component of the geodesic equation,
\begin{equation}
    \frac{\d^2\eta}{\d\hat\lambda^2} + 2(\frac{\d\eta}{\d\hat\lambda})^2\frac{\d\Psi}{\d\eta}+2\frac{\d\eta}{\d\hat\lambda}\frac{\d x^i}{\d\hat\lambda}\frac{\partial\Psi}{\partial x^i} = 0,
\end{equation}
where the derivative $\d\Psi/\d\eta = \partial_\eta\Psi+(\d x^i/\d\eta)\partial_i\Psi$ is along the perturbed ray. This allows us to rewrite the other (spatial) geodesic equations as
\begin{equation}
    \frac{\d^2x^i}{\d\eta^2} - 2\frac{\d x^i}{\d\eta}(\frac{\d\Psi}{\d\eta} + \frac{\d x^j}{\d\eta}\frac{\partial\Psi}{\partial x^j})+2\gamma^{ij}\frac{\partial\Psi}{\partial x^j}+\bar\Gamma^i_{jk}\frac{\d x^j}{\d\eta}\frac{\d x^k}{\d \eta} = 0,
\end{equation}
where $\bar\Gamma^i_{jk}$ are the connection coefficients of the unperturbed three geometry $\gamma_{ij}$. Without loss of generality we consider an observer located at the origin of the spatial coordinates, meaning we are interested in rays that end at $x^i = 0$. In that case $\d \chi/\d \eta = -1 + O(\Psi)$, $\d \theta/\d\eta = O(\Psi)$ and $\d \phi/\d\eta = O(\Psi)$ (with $\theta$ and $\phi$ angular spatial coordinates). For the case of such rays we can evaluate the connection coefficients and rewrite the spatial geodesic equations as
\begin{gather}
    \frac{\partial^2\chi}{\d\eta^2}+2\frac{\d\Psi}{\d\eta}=0,\\
    \frac{\d^2\theta}{\d\eta^2}-2\frac{\d\ln f_K(\chi)}{\d\chi}\frac{\d\theta}{\d\eta}+\frac{2}{f^2_K(\chi)}\frac{\partial\Psi}{\partial\theta}=0,\\
    \frac{\d^2\phi}{\d\eta^2} - 2\frac{\d \ln f_K(\chi)}{\d\chi}\frac{\d\phi}{\d\eta}+\frac{2}{f^2_K(\chi)}\frac{1}{\sin^2(\theta)}\frac{\partial\Psi}{\partial\phi}=0.
\end{gather}
Integrating twice and using the null condition we can rewrite the the first equation as
\begin{equation}
    \chi = \eta_0 - \eta - 2\int_{\eta_0}^\eta\Psi\d\eta',
\end{equation}
where the integral is along the ray. Since we are working at first order in $\Psi$ we can evaluate the integral along the unperturbed path, this is known as the \textbf{Born approximation}.\textit{Next follows a short discussion of why perturbations in observation time can be ignored, include?} Doing the same for the other two equations gives
% TODO: make n a proper vector (bold)
\begin{gather}
    \theta(\eta_0 - \chi_*)=\theta_0 - \int_0^{\chi_*}\d\chi\frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}2\frac{\partial}{\partial \theta}\Psi(\chi\hat n, \eta_0 - \chi),\\
    \phi(\eta_0 - \chi_*)=\theta_0 - \int_0^{\chi_*}\d\chi\frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}\frac{2}{\sin^2\theta}\frac{\partial}{\partial \phi}\Psi(\chi\hat n, \eta_0 - \chi),
\end{gather}
where $\theta_0$ and $\phi_0$ label the line of sight $\hat n$. The displacement vector $\mathbf{\alpha}$ indicates the infinitesimal displacement of the lensed rays from their unpetrubed observed angles and is related to $\theta$ and $\phi$ through
$$
\alpha_\theta = \theta - \theta_0, \quad \alpha_\phi = \sin^2\theta(\phi - \phi_0).
$$
This allows us to summarize the above results as
\begin{equation}
    \mathbf \alpha = -2 \int_0^{\chi_*}\d \chi \frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}\nabla_{\hat{\mathbf n}}\Psi(\chi\hat{\mathbf n};\eta_0 - \chi).
\end{equation}
When working in the weak field limit (i.e. small spacetime curvature) the $00$ component of Einsteins equations give us a poisson equation \cite{dodelson2020modern},
\begin{equation}
    -\nabla^2\Psi(\textbf{x}, \eta) =  4\pi G a(\eta)^2\delta\rho(\mathbf x, \eta)
\end{equation}
where $K$ is $0$, $1$, or $-1$ depending on the curvature of the universe, $\Delta = \partial_x^2 + \partial_y^2 + \partial_z^2$ is the 3 dimensional laplace operator, $\delta\bar\rho$ is the comoving total energy perturbation (evaluated in the rest-frame of the total energy). %TODO: proper definition for anisotropic stress and explain why we can ignore it

@@

\subsection{Regarding window functions}
The window function for light emitted from a source at fixed radius $\chi_*$ is $(\chi-\chi^*)/(\chi_*\chi)$, if the source is instead distributed over the radius as $p(\chi)$, with $p(\chi)$ normalized to integrate to 1, we generalize the window function as
\begin{equation*}
    W(\chi) = \int_\chi^{\chi_*}\d\chi' p(\chi')\frac{\chi'-\chi}{\chi'\chi}.
\end{equation*}
The integration limit is up to surface of last scattering because further than that no source can contribute to the observed image. In the case of an image of the CMB we can take $p(\chi')=\delta(\chi'-\chi_*)$, in which case the window function reduces back to $(\chi_*-\chi)/(\chi_*\chi)$. In the case of different sources, we will differentiate between their respective window functions with a subscript. In this case the factor that changes is the distribution function. The sources considered in this paper are only the CMB and selections of galaxies from galaxy surveys.

\section{Weak Lensing Statistics}\label{sec:weaklensstats}

@@

\section{Calculation Details}\label{sec:calcdetails}
\subsection{Numerical Considerations}
% I used CAMB, how I took derivative, interpolation, fisher matrix sampling, matter bispectrum in terms of power spectrum

Running cosmological simulations to obtain the matter power spectra and other functions was done with the CAMB package (REF).

The derivatives of the lensing spectra were done numerically using a four point central difference formula, given by
\begin{equation*}
    f'(x) = \frac{-f(x + 2h) + 8f(x+h) - 8 f(x - h) + f(x - 2h)}{12h} + \mathcal O (h^4).
\end{equation*}
$h$ was chosen to be $2.5\%$ of the $x$ value. Testing showed this to generally be the best choise to avoid both the inacuracies from numerical errors (relevant for small $h$) and inacuracies due to the $\mathcal O(h^4)$ error inherent in the four point finite difference formula. The derivative w.r.t. $\sigma_8$ was an exception as it could not be taken by varying it directly and so was calculated implicitly using the chain rule.

To speed up the calculation of the lensing powerspectra and especially the bispectra, an interpolation approach was used. Due to being 3 dimensional, it was not feasible to interpolate the lensing bispectrum directly. We instead chose to interpolate all 1 and 2 dimensional functions that the bispectra are based on and then calculate the bispectra with the interpolated functions instead. This significantly sped up the computation at the cost of small errors in the values in the bispectrum ($\sim 1\%$) when compared with the uninterpolated values.

>>>>

% !TEX root = /Users/jonasfrugte/Desktop/ResearchProject/paper.tex
\documentclass[11pt]{article} % Choose a document class, like 'article', 'report', or 'revtex4-2' for scientific papers

% Packages to enhance functionality
\usepackage[utf8]{inputenc}   % Allows UTF-8 character encoding
\usepackage{amsmath}          % For advanced math typesetting
\usepackage{amssymb}          % Additional math symbols
\usepackage{graphicx}         % For including images
\usepackage{hyperref}         % For hyperlinks in the document
\usepackage{cite}             % For handling references better
\usepackage{geometry}         % For setting up page geometry
\geometry{margin=1in}         % 1-inch margins all around
\usepackage{authblk}          % For author affiliations
\usepackage{siunitx}          % For consistent typesetting of units
\usepackage{subcaption}       % For subfigures within a figure
\usepackage{float}            % To control float positioning (e.g., figures/tables)
\usepackage{enumitem}         % Better control over list formatting
\usepackage{tocloft}
\usepackage{tikz}
\usepackage{tikz-3dplot}
\usetikzlibrary{shapes.geometric, arrows, backgrounds}
\usepackage{empheq}
\usepackage{booktabs}

@@

@@

\begin{document}

\textbf{
    \Huge \noindent 
    Parameter Forecasts from Convergence $\times$ Shear Bispectra 
    \vspace{0.5cm}
    \\
    \normalsize
    Jonas Frugte, Daan Meerburg\\
    24 October 2024
    }\\
\vspace{0.5cm}
\hrule

\begin{abstract}
    abstract text
\end{abstract}
\vspace{0.5cm}
\hrule

\tableofcontents
\pagebreak
\section{Conventions}
Use $\mathbf k$ for wave numbers (Fourier space), $l$ for multipole moments, $m$ for magnetic quantum number. $P^{XY}(k, \eta, \eta')$ and $B^{XYZ}(k_1,k_2,k_3, \eta_1, \eta_2, \eta_3)$ correspond to bi and power spectra of fields $X$, $Y$, and $Z$ in Fourier space. $P_l^{XY}(\eta,\eta')$ and $B_{l_1l_2l_3}^{XYZ}(\eta_1, \eta_2, \eta_3)$ correspond to the multipole moments of the bi and power spectra of fields $X$, $Y$, and $Z$. If all times are the same we may instead write only one time argument and if the time is current time then we omit the time argument entirely. In other words,
$$
P_l^{XY}(\eta):=P_l^{XY}(\eta,\eta), \quad P_l^{XY}:=P_l^{XY}(\eta_{\text{current}}), \quad \text{similarly for all other spectra}.
$$
If both fields are the same, we may only denote one field, e.g. 
$$
P_l^{X}(\eta) := P_l^{XX}(\eta).
$$

Our Fourier convention is
\begin{align}
    &\text{Forward Transform:} & \quad \tilde{f}(k) &= \int_{-\infty}^{\infty} f(x) \, e^{-i k x} \, dx\\
    &\text{Inverse Transform:} & \quad f(x) &= \frac{1}{2\pi} \int_{-\infty}^{\infty} \tilde{f}(k) \, e^{i k x} \, dk
\end{align}
we choose this convention because it is in line with the convention used in the CAMB source code. We work in natural units where $c=1$.

@@

\begin{figure}[h!]
    \centering
    \input{figures/calc_flowchart.tikz}
    \caption{Flowchart of steps involved in calculating the minimum error/uncertainty in cosmological parameter estimates using convergence and shear power- and bi-spectra.}
\end{figure}

\section{Weak Lensing}
\subsection{Approximations Made / Regime}
\subsection{Perturbed Photon Paths}
Work in \textbf{conformal newtonian gauge}. In this section we'll work in units where $c=1$. Denoting conformal time and conformal radial distance by $\eta$ and $\chi$, respectively, the perturbed line element is given by
\begin{equation}
    \d s^2 = a^2(\eta)((1 + 2\Psi_N) \d \eta^2 - (1 + 2\Phi_N) \gamma_{ij}\d x^i \d x^j) 
\end{equation}

@@

In the unperturbed there are incoming radial solutions of the form $\chi = \eta_0 - \eta$. We can express $\hat\lambda$ in terms of $\eta$ using The 0-component of the geodesic equation,
    \frac{\d^2\eta}{\d\hat\lambda^2} + 2(\frac{\d\eta}{\d\hat\lambda})^2\frac{\d\Psi}{\d\eta}+2\frac{\d\eta}{\d\hat\lambda}\frac{\d x^i}{\d\hat\lambda}\frac{\partial\Psi}{\partial x^i} = 0,
where the derivative $\d\Psi/\d\eta = \partial_\eta\Psi+(\d x^i/\d\eta)\partial_i\Psi$ is along the perturbed ray. This allows us to rewrite the other (spatial) geodesic equations as
    \frac{\d^2x^i}{\d\eta^2} - 2\frac{\d x^i}{\d\eta}(\frac{\d\Psi}{\d\eta} + \frac{\d x^j}{\d\eta}\frac{\partial\Psi}{\partial x^j})+2\gamma^{ij}\frac{\partial\Psi}{\partial x^j}+\bar\Gamma^i_{jk}\frac{\d x^j}{\d\eta}\frac{\d x^k}{\d \eta} = 0,
where $\bar\Gamma^i_{jk}$ are the connection coefficients of the unperturbed three geometry $\gamma_{ij}$. Without loss of generality we consider an observer located at the origin of the spatial coordinates, meaning we are interested in rays that end at $x^i = 0$. In that case $\d \chi/\d \eta = -1 + O(\Psi)$, $\d \theta/\d\eta = O(\Psi)$ and $\d \phi/\d\eta = O(\Psi)$ (with $\theta$ and $\phi$ angular spatial coordinates). For the case of such rays we can evaluate the connection coefficients and rewrite the spatial geodesic equations as
    \frac{\partial^2\chi}{\d\eta^2}+2\frac{\d\Psi}{\d\eta}=0,\\
    \frac{\d^2\theta}{\d\eta^2}-2\frac{\d\ln f_K(\chi)}{\d\chi}\frac{\d\theta}{\d\eta}+\frac{2}{f^2_K(\chi)}\frac{\partial\Psi}{\partial\theta}=0,\\
    \frac{\d^2\phi}{\d\eta^2} - 2\frac{\d \ln f_K(\chi)}{\d\chi}\frac{\d\phi}{\d\eta}+\frac{2}{f^2_K(\chi)}\frac{1}{\sin^2(\theta)}\frac{\partial\Psi}{\partial\phi}=0.
Integrating twice and using the null condition we can rewrite the the first equation as
    \chi = \eta_0 - \eta - 2\int_{\eta_0}^\eta\Psi\d\eta',
where the integral is along the ray. Since we are working at first order in $\Psi$ we can evaluate the integral along the unperturbed path, this is known as the \textbf{Born approximation}.\textit{Next follows a short discussion of why perturbations in observation time can be ignored, include?} Doing the same for the other two equations gives
% TODO: make n a proper vector (bold)
    \theta(\eta_0 - \chi_*)=\theta_0 - \int_0^{\chi_*}\d\chi\frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}2\frac{\partial}{\partial \theta}\Psi(\chi\hat n, \eta_0 - \chi),\\
    \phi(\eta_0 - \chi_*)=\theta_0 - \int_0^{\chi_*}\d\chi\frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}\frac{2}{\sin^2\theta}\frac{\partial}{\partial \phi}\Psi(\chi\hat n, \eta_0 - \chi),
where $\theta_0$ and $\phi_0$ label the line of sight $\hat n$. The displacement vector $\mathbf{\alpha}$ indicates the infinitesimal displacement of the lensed rays from their unpetrubed observed angles and is related to $\theta$ and $\phi$ through
$$
\alpha_\theta = \theta - \theta_0, \quad \alpha_\phi = \sin^2\theta(\phi - \phi_0).
$$
This allows us to summarize the above results as
    \mathbf \alpha = -2 \int_0^{\chi_*}\d \chi \frac{f_K(\chi_* - \chi)}{f_K(\chi_*)f_K(\chi)}\nabla_{\hat{\mathbf n}}\Psi(\chi\hat{\mathbf n};\eta_0 - \chi).
When working in the weak field limit (i.e. small spacetime curvature) the $00$ component of Einsteins equations give us a poisson equation \cite{dodelson_modern_cosmology},
    -\nabla^2\Psi(\textbf{x}, \eta) =  4\pi G a(\eta)^2\delta\rho(\mathbf x, \eta)
where $K$ is $0$, $1$, or $-1$ depending on the curvature of the universe, $\Delta = \partial_x^2 + \partial_y^2 + \partial_z^2$ is the 3 dimensional laplace operator, $\delta\bar\rho$ is the comoving total energy perturbation (evaluated in the rest-frame of the total energy). %TODO: proper definition for anisotropic stress and explain why we can ignore it

@@

\subsection{Regarding window functions}
The window function for light emitted from a source at fixed radius $\chi_*$ is $(\chi-\chi^*)/(\chi_*\chi)$, if the source is instead distributed over the radius as $p(\chi)$, with $p(\chi)$ normalized to integrate to 1, we generalize the window function as
\begin{equation*}
    W(\chi) = \int_0^{\chi_*}\d\chi' p(\chi')\frac{\chi'-\chi}{\chi'\chi}.
\end{equation*}
The integration limit is up to surface of last scattering because further than that no source can contribute to the observed image. In the case of an image of the CMB we can take $p(\chi')=\delta(\chi'-\chi_*)$, in which case the window function reduces back to $(\chi_*-\chi)/(\chi_*\chi)$. In the case of different sources, we will differentiate between their respective window functions with a subscript. In this case the factor that changes is the distribution function. The sources considered in this paper are only the CMB and selections of galaxies from galaxy surveys.

\section{Spherical Harmonics}
\subsection{definition}
The \textbf{spherical harmonics} $Y_{l}^{m}(\theta, \phi)$ with $l \geq 0$ and $-l\leq m\leq l$ are defined on a sphere in 3D space ($0\leq\theta\leq \pi$ and $0\leq\phi<2\pi$, also denoted $S^2$). They arise naturally when solutions of the laplace, helmholtz, and schrodinger equations in 3D space are decomposed into radial and angular parts, $f(r, \theta, \phi) = f_{\text{spher}}(r)f_{\text{angular}}(\theta, \phi)$. They can be written explicitely as
\begin{equation}
    Y^m_l(\theta, \phi) := \sqrt{\frac{2l+1}{4\pi}\frac{(l-m)!}{(l+m)!}}P_l^m(\cos\theta)e^{im\phi},
\end{equation}
where $P_l^m$ are \textbf{associated legendre functions}. They are defined in terms of the \textbf{legendre functions} as
\begin{equation}
    P_l^m(x) = \begin{cases}
        (1-x^2)^{m/2}\frac{\d^m}{\d x^m}P_l(x), & m \geq 0,\\
        (-1)^m\frac{(l+m)!}{(l-m)!}P_l^{-m}(x), & m < 0.
    \end{cases}
\end{equation}
Finally, the legendre functions themselves are implicitly defined through the generating function
% TODO: find source
\begin{equation}
    G(x, t) = \frac{1}{\sqrt{1 - 2xt + t^2}} = \sum_{l=0}^\infty P_l(x)t^l \implies P_l(x) = \frac{1}{l!}\frac{\d ^l}{\d t^l}G(x, 0).
\end{equation}
Here we require $x$ to be a real number between $-1$ and $1$ and $t$ a complex number with magnitude less than 1.

\subsection{orthonormality}
Defining the inner product
\begin{equation}
    \langle f(\theta, \phi)|g(\theta, \phi)\rangle := \int_0^{2\pi}\d\phi\int_0^\pi\sin\theta\d\theta f(\theta, \phi)^*g(\theta, \phi)
\end{equation}
on the space of functions from $S^2$ to $\mathbb C$, we can show that the spherical harmonics are orthonormal, i.e.
\begin{equation}
    \langle Y^{m'}_{l'} | Y^{m}_{l} \rangle = \delta_{mm'}\delta_{ll'}.
\end{equation}

We first show orthogonality of the associated legendre functions. They satisfy the differential equation
\begin{equation}
    \frac{\d}{\d x}((1-x^2)\frac{\d P^m_l(x)}{\d x}) + (l(l + 1) - \frac{m^2}{1 - x^2})P^m_l(x) = 0,
\end{equation}
on the interval $[-1, 1]$, which is of \textbf{Sturm-Liouville form},
\begin{equation}
    \frac{\d}{\d x}(p(x)\frac{\d y}{\d x}) + (-\lambda - q(x))y = 0.
\end{equation}
$\lambda$ is refered to as the eigenvalue of $y$ in this case. Some texts generalize the $\lambda$ term to $\lambda w(x)$, but we will not need these extra complications here. %TODO: find a ref
It is a general result for $y_n$ and $y_m$ with distinct eigenvalues $\lambda_n$ and $\lambda_m$ that they are orthogonal. Consider the space of real functions on (wlog) $[-1, 1]$. We can define an inner product
\begin{equation}
    \langle f | g \rangle := \int_{-1}^1 f(x)g(x)\d x.
\end{equation}
In this case the sturm Liouville problem can be restated as finding the eigenvalues of a differential operator $\mathcal L$, i.e.
\begin{equation}
\mathcal L [y] = \lambda, \quad \mathcal L[y] := \frac{\d}{\d x}(p(x)\frac{\d y}{\d x}) - q(x)y.
\end{equation}
It can then be easily verified that $\mathcal L$ is a hermitian operator, i.e. $\langle \mathcal L f | g \rangle = \langle f | \mathcal L g \rangle $ if $p(x)$ vanishes on the boundaries. % TODO: verify the boundary part
It is then a standard result that eigenvalues of the hermitian operator with distinct eigenvalues are orthogonal. Given that the $P^m_l$ are eigenfunctions of a sturm liouvilly problem with associated eigenvalues $l(l+1)$, they must be orthogonal. In the case that they are equal, we the integral can be analytically calculated, giving the result
\begin{equation}
    \int_{-1}^1 P^m_l(x)P^m_{l'}(x) \d x = \frac{2(l+m)!}{(2l+1)(l-m)!}\delta_{ll'}.
\end{equation}
Orthonormality of the spherical harmonics is than checked by using the definition in terms of the associated legendre functions. Define the normalisation constant of $Y_m^l$ as $N^m_l$, then
\begin{gather}
    \int \d\theta\d \phi \sin\theta (Y^m_l)^* Y^{m'}_{l'} = N_l^mN_{l'}^{m'} \int_0^{2\pi} e^{-im\phi}e^{im'\phi}\int_0^\pi P^m_l(\cos\theta)P^{m'}_{l'}(\cos\theta)\sin\theta\d\theta \\
    = (N_l^m)^2 \delta_{mm'}\frac{2(l+m)!}{(2l+1)(l-m)!}\delta_{ll'} = \delta_{mm'}\delta_{ll'}.
\end{gather}

\subsection{completenes}
Take an arbitrary function $f(\theta, \phi)$. Define harmonic coefficients $c_{lm} = \langle Y_l^m | f \rangle$, then we aim to prove that, for continuous $f$, $f(\theta, \phi) = \sum_{lm} c_{lm}Y^m_l(\theta\phi)$. We first require 2 intermediate results. The legendre addition formula is a well known resulting stating that
\begin{equation}
    P_l(\cos\gamma) = \frac{4\pi}{2l + 1} \sum_{-l\leq m \leq l} Y^m_l(\theta', \phi')^* Y^m_l(\theta, \phi),
\end{equation}
where $\gamma$ is the angle between the vectors pointing in the directions $(\theta', \phi')$ and $(\theta, \phi)$. Additionally, in E W Thompson % TODO: get proper ref
it is shown that, for points at which $f$ is continuous,
\begin{equation}
   f(\theta, \phi) = \sum_{l\geq 0} \frac{2l + 1}{4\pi} \int^\pi_0\int^{2\pi}_{0}f(\theta',\phi')P_l(\cos\gamma)\sin\theta'\d\theta'\d\phi'.
\end{equation}
Putting these two results together we write 
\begin{gather}
    \sum_{lm} c_{lm}Y^m_l(\cos\theta\phi) = \sum_{lm} \int \d\phi'\d\theta'\sin\theta'Y^m_l(\theta',\phi')^*f(\theta',\phi')Y^m_l(\theta,\phi) \\
    = \sum_l \int \d\theta'\d\phi'\sin\theta' P_l(\cos\gamma)\frac{2l+1}{4\pi}f(\theta',\phi') = f(\theta, \phi).
\end{gather}
Thus the spherical harmonics form a complete orthonormal basis of the real functions on $S^2$

\section{Weak Lensing Statistics}

@@

\section{Cosmological Perturbation Theory}
